---
title: "Predicting Football Results: Statistical Learnings Methods"
author: "LABARTHE, S."
output: pdf_document
date: "2026"

---

# CODEBOOK

## Identification 

* `match_id` - Unique numeric identifier for each match
* `date` - Date of the match (YYYY-MM-DD format)
* `year` - Year of the match (extracted from date)
* `country_name` - Name of the focal team
* `opponent_name` - Name of the opposing team
* `dyad_name` - Alphabetically sorted team pair identifier (e.g., "England-Germany")

## Football variables

* `home_result` - Match result from focal team's perspective (1=win, 0.5=draw, 0=loss)
* `away_result` - Match result from opponent's perspective (1=win, 0.5=draw, 0=loss)
* `goals_for` - Number of goals scored by focal team
* `goals_against` - Number of goals conceded by focal team
* `goal_difference` - Goals for minus goals against
* `margin` - Absolute goal difference ($\vert$goals_for - goals_against$\vert$)
* `home` - Home advantage indicator (1=home team, -1=away team, 0=neutral)
* `hga` - Home ground advantage (100 if not neutral, 0 if neutral)
* `tournament` - Name of the tournament/competition
* `tournament_weight` - Importance weight assigned to tournament (60=World Cup, 50=continental cups, 40=qualifiers, 20=friendly)
* `goal_differ_multiplier` - Multiplier based on margin of victory (1 for $\leq$ 1 goal, up to 1.75+ for blowouts)
* `k` - Weighted importance score (tournament_weight * goal_differ_multiplier)
* `k2` - Time-decayed importance score (k adjusted by years since match)
* `city` - City where the match was played
* `country` - Country where the match was played

* `country_elo` - Focal team's Elo rating before the match
* `opponent_elo` - Opponent's Elo rating before the match
* `elo_diff` - Elo rating difference (country_elo - opponent_elo)
* `home_update` - Elo update value for home team after match
* `away_update` - Elo update value for away team after match

* `total_matches` - Cumulative number of matches played by focal team up to this point
* `wins` - Cumulative number of wins by focal team
* `draws` - Cumulative number of draws by focal team
* `losses` - Cumulative number of losses by focal team
* `win_share` - Proportion of matches won (wins / total_matches)
* `win_prob_naive` - Naive win probability including draws as half-wins ((wins + 0.5*draws) / total_matches)
* `last5_result_avg` - Average match result over last 5 matches (1=win, 0.5=draw, 0=loss)
* `last5_result_sum` - Sum of match results over last 5 matches (max 5)
* `last_year_goals_for_avg` - Average goals scored per match in last 365 days
* `last_year_goals_against_avg` - Average goals conceded per match in last 365 days
* `last_3months_goals_for_avg` - Average goals scored per match in last 90 days
* `last_3months_goals_against_avg` - Average goals conceded per match in last 90 days

* `h2h_matches` - Cumulative number of matches played against this specific opponent
* `h2h_wins` - Cumulative number of wins against this specific opponent
* `h2h_draws` - Cumulative number of draws against this specific opponent
* `h2h_losses` - Cumulative number of losses against this specific opponent
* `h2h_win_rate` - Win rate against this specific opponent (h2h_wins / h2h_matches)
* `h2h_points_per_game` - Average points per game against opponent (wins=3, draws=1)
* `last5_vs_opponent_result_avg` - Average match result over last 5 matches against this specific opponent
* `last5_vs_opponent_goals_for_avg` - Average goals scored over last 5 matches against this specific opponent
* `last5_vs_opponent_goals_against_avg` - Average goals conceded over last 5 matches against this specific opponent

* `opponent_home_result` - Match result from opponent's perspective
* `opponent_goals_for` - Goals scored by opponent
* `opponent_goals_against` - Goals conceded by opponent
* `opponent_last5_result_avg` - Opponent's average match result over last 5 matches
* `opponent_last_year_goals_for_avg` - Opponent's average goals scored per match in last 365 days
* `opponent_last_year_goals_against_avg` - Opponent's average goals conceded per match in last 365 days
* `opponent_win_share` - Opponent's proportion of matches won historically

## Geopolitical variables 

* `bicc_milper` - Military Personnel Index (BICC)
* `bicc_milexp` - Military Expenditure Index (BICC)
* `wdi_expmil` - Military expenditure as % of GDP (World Bank)
* `wdi_pop` - Total population (World Bank)
* `wdi_popurb` - Urban population (World Bank)
* `wdi_birth` - Birth rate per 1,000 people (World Bank)
* `wdi_death` - Death rate per 1,000 people (World Bank)
* `wdi_mortinf` - Infant mortality per 1,000 live births (World Bank)
* `wdi_lifexp` - Life expectancy at birth (World Bank)
* `wpp_medianage` - Median age as of July 1 (UN World Population Prospects)
* `wpp_sexratio` - Sex ratio (males per 100 females) (UN WPP)
* `wdi_export` - Exports of goods and services as % of GDP (World Bank)
* `wdi_import` - Imports of goods and services as % of GDP (World Bank)
* `gle_exp` - Total exports in millions of current USD (Gleditsch)
* `gle_imp` - Total imports in millions of current USD (Gleditsch)
* `wdi_area` - Land area in square kilometers (World Bank)
* `wdi_eneuse` - Energy use in kg of oil equivalent per capita (World Bank)
* `vdem_polyarchy` - Democracy level index (V-Dem)
* `ucdp_type3` - Number of internal armed conflicts (UCDP)

* `opponent_bicc_milper` - Opponent's Military Personnel Index
* `opponent_bicc_milexp` - Opponent's Military Expenditure Index
* `opponent_wdi_expmil` - Opponent's military expenditure as % of GDP
* `opponent_wdi_pop` - Opponent's total population
* `opponent_wdi_popurb` - Opponent's urban population
* `opponent_wdi_birth` - Opponent's birth rate per 1,000
* `opponent_wdi_death` - Opponent's death rate per 1,000
* `opponent_wdi_mortinf` - Opponent's infant mortality per 1,000
* `opponent_wdi_lifexp` - Opponent's life expectancy at birth
* `opponent_wpp_medianage` - Opponent's median age
* `opponent_wpp_sexratio` - Opponent's sex ratio
* `opponent_wdi_export` - Opponent's exports as % of GDP
* `opponent_wdi_import` - Opponent's imports as % of GDP
* `opponent_gle_exp` - Opponent's total exports in millions USD
* `opponent_gle_imp` - Opponent's total imports in millions USD
* `opponent_wdi_area` - Opponent's land area in sq km
* `opponent_wdi_eneuse` - Opponent's energy use per capita
* `opponent_vdem_polyarchy` - Opponent's democracy level
* `opponent_ucdp_type3` - Opponent's number of internal armed conflicts

* `milper_diff` - Difference in military personnel (focal - opponent)
* `milexp_diff` - Difference in military expenditure index (focal - opponent)
* `milexp_gdp_diff` - Difference in military expenditure % of GDP (focal - opponent)
* `pop_ratio` - Log ratio of total population (log(focal_pop / opponent_pop))
* `popurb_ratio` - Log ratio of urban population (log(focal_urban / opponent_urban))
* `popurb_pct_diff` - Difference in urban population percentage
* `birth_rate_diff` - Difference in birth rate (focal - opponent)
* `death_rate_diff` - Difference in death rate (focal - opponent)
* `infant_mort_diff` - Difference in infant mortality (focal - opponent)
* `life_exp_diff` - Difference in life expectancy (focal - opponent)
* `median_age_diff` - Difference in median age (focal - opponent)
* `sex_ratio_diff` - Difference in sex ratio (focal - opponent)
* `export_gdp_diff` - Difference in exports % of GDP (focal - opponent)
* `import_gdp_diff` - Difference in imports % of GDP (focal - opponent)
* `total_export_ratio` - Log ratio of total exports (log(focal_exports / opponent_exports))
* `total_import_ratio` - Log ratio of total imports (log(focal_imports / opponent_imports))
* `trade_openness_diff` - Difference in trade openness (exports+imports % of GDP)
* `area_ratio` - Log ratio of land area (log(focal_area / opponent_area))
* `energy_use_diff` - Difference in energy use per capita (focal - opponent)
* `democracy_diff` - Difference in democracy level (focal - opponent)
* `conflict_diff` - Difference in number of internal conflicts (focal - opponent)
* `conflict_any` - Whether at least one team has internal conflict (Yes/No)
* `home_advantage_geo` - Composite geopolitical home advantage score
* `geo_data_complete` - Whether all geopolitical variables are available (TRUE/FALSE)
* `geo_any_missing` - Whether any geopolitical data is missing (TRUE/FALSE)

```{r, echo = F, message = F, warning = F}
library(dplyr)
library(elo)
library(formattable)
library(lubridate)
library(readr)
library(rvest)
library(showtext)
library(tidyverse)
library(glue)
```

# DATA
## Next match scrap
```{r, echo = F, message = F, warning = F}
# Pick URL
url_futur <- "https://www.programmetv-foot.fr/calendrier/matchs-internationnaux"

# Read the HTML page
page <- read_html(url_futur)
page

# Extract the Data
matches <- page |>
  html_elements("div.div_match")

home_team <- matches |> 
  html_elements(".div_dom .club") |> 
  html_text2()

away_team <- matches |> 
  html_elements(".div_ext .club") |> 
  html_text2()

datetime  <- matches |> 
  html_elements("time") |> 
  html_attr("datetime")

# Build Data Frame
df <- tibble(
  home_team = home_team,
  away_team = away_team,
  date      = as.Date(datetime)
)

df
```

## Foot data 
```{r, echo = F, message = F, warning = F}
rm(list = ls())

url_past  <- "https://raw.githubusercontent.com/martj42/international_results/master/results.csv"
dyadic_data <- read_csv(url_past)

dyadic_data <- dyadic_data %>%
  mutate(
    home_result = case_when(
      home_score >  away_score ~ 1,
      home_score == away_score ~ 0.5,
      home_score <  away_score ~ 0
    ),
    away_result = case_when(
      home_score <  away_score ~ 1,
      home_score == away_score ~ 0.5,
      home_score >  away_score ~ 0
    ),
    margin = abs(home_score - away_score)
  ) %>%
  drop_na(margin)

dyadic_data <- dyadic_data %>%
  mutate(
    hga = 100 * !neutral,

    tournament_weight = case_when(
      tournament == "FIFA World Cup"                                    ~ 60,
      tournament %in% c("Confederations Cup", "African Cup of Nations",
                        "Copa America", "UEFA Euro", "AFC Asian Cup")   ~ 50,
      str_detect(tolower(tournament), "qualification")                  ~ 40,
      str_detect(tolower(tournament), " cup")                           ~ 40,
      tournament %in% c("UEFA Nations League","CONCACAF Nations League")~ 40,
      tournament == "Friendly"                                          ~ 20,
      TRUE                                                              ~ 30
    ),

    goal_differ_multiplier = case_when(
      margin <= 1 ~ 1,
      margin == 2 ~ 1.5,
      margin == 3 ~ 1.75,
      margin >= 4 ~ 1.75 + (margin - 3) / 8
    ),

    k        = tournament_weight * goal_differ_multiplier,
    k2       = tournament_weight * goal_differ_multiplier /
                 0.1 * (year(today()) - year(date)) + 1,
    match_id = row_number()
  ) %>%
  relocate(match_id)

elo_model <- elo.run(
  data = dyadic_data,
  formula = home_result ~ adjust(home_team, hga) + away_team + k(k)
)

dyadic_data <- elo_model %>%
  as_tibble() %>%
  mutate(match_id = row_number()) %>%
  select(
    match_id,
    home_update    = update.A,
    away_update    = update.B,
    home_elo_after = elo.A,
    away_elo_after = elo.B
  ) %>%
  mutate(
    elo_a    = home_elo_after - home_update,
    elo_b    = away_elo_after - away_update,
    elo_diff_dyadic = elo_a - elo_b
  ) %>%
  right_join(dyadic_data, by = "match_id")

elo_model_lm <- lm(home_result ~ elo_diff_dyadic, data = dyadic_data)

home_team_matches <- dyadic_data %>%
  rename(
    country_name = home_team,
    opponent_name = away_team,
    goals_for = home_score,
    goals_against = away_score,
    country_elo = elo_a, 
    opponent_elo = elo_b, 
  ) %>% 
  mutate(
    # Home team perspective
    home_result = case_when(
      goals_for > goals_against ~ 1,
      goals_for == goals_against ~ 0.5,
      goals_for < goals_against ~ 0
    ),
    away_result = case_when(
      goals_for < goals_against ~ 1,
      goals_for == goals_against ~ 0.5,
      goals_for > goals_against ~ 0
    ),
    elo_diff = country_elo - opponent_elo,
    goal_difference = goals_for - goals_against, 
    home = ifelse(neutral == FALSE, 1, 0), # 1 for advantage
    year = as.numeric(format(date, "%Y")),     
    dyad_name = paste0(pmin(country_name, opponent_name), "-", 
                       pmax(country_name, opponent_name))
  )

away_team_matches <- dyadic_data %>%
  rename(
    country_name = away_team,
    opponent_name = home_team,
    goals_for = away_score,
    goals_against = home_score,
    country_elo = elo_b, 
    opponent_elo = elo_a, 
  ) %>% 
  mutate(
    # Away team perspective (recalculated!)
    home_result = case_when(
      goals_for > goals_against ~ 1,   # away team scored more = win
      goals_for == goals_against ~ 0.5,
      goals_for < goals_against ~ 0
    ),
    away_result = case_when(
      goals_for < goals_against ~ 1,
      goals_for == goals_against ~ 0.5,
      goals_for > goals_against ~ 0
    ),
    elo_diff = country_elo - opponent_elo,
    goal_difference = goals_for - goals_against, 
    home = ifelse(neutral == FALSE, -1, 0), # -1 for disadvantage
    year = as.numeric(format(date, "%Y")),     
    dyad_name = paste0(pmin(country_name, opponent_name), "-", 
                       pmax(country_name, opponent_name))
  )

monadic_matches <- bind_rows(home_team_matches, 
                             away_team_matches) %>% 
  select(
    country_name,           # 1st column
    year,     
    date, 
    opponent_name,          # 3rd column
    dyad_name,              # 4th column
    home_result, 
    goals_for, 
    goals_against, 
    everything(), 
    -neutral, 
    -elo_diff_dyadic, 
    -home_elo_after,
    -away_elo_after) %>%
  arrange(country_name, 
          desc(year), 
          desc(date))

# Check for inconsistencies
inconsistent <- monadic_matches %>%
  mutate(
    should_be_win = case_when(
      goals_for > goals_against ~ 1,
      goals_for == goals_against ~ 0.5,
      goals_for < goals_against ~ 0
    ),
    is_inconsistent = (home_result != should_be_win)
  ) %>%
  filter(is_inconsistent == TRUE)

# Should be 0 rows
nrow(inconsistent)

# If any inconsistencies remain, show them
if(nrow(inconsistent) > 0) {
  print(inconsistent %>% 
          select(country_name, opponent_name, goals_for, goals_against, home_result, should_be_win))
}

library(zoo)  # for rolling calculations if needed

monadic_data <- monadic_matches %>% 
  arrange(country_name, desc(date)) %>%  # ensure chronological order per team
  
  # Group by country to create team-level rolling statistics
  group_by(country_name) %>%
  mutate(
    # 1. Aggregate home_result of last 5 matches (all matches, not just home)
    last5_result_avg = rollapplyr(home_result, width = 5, 
                                   FUN = mean, fill = NA, align = "right"),
    last5_result_sum = rollapplyr(home_result, width = 5, 
                                   FUN = sum, fill = NA, align = "right"),
    
    # 2. Average goals for/against last year (365 days)
    last_year_goals_for_avg = map_dbl(date, ~ {
      matches_last_year = goals_for[between(date, .x - 365, .x - 1)]
      if(length(matches_last_year) > 0) mean(matches_last_year) else NA
    }),
    last_year_goals_against_avg = map_dbl(date, ~ {
      matches_last_year = goals_against[between(date, .x - 365, .x - 1)]
      if(length(matches_last_year) > 0) mean(matches_last_year) else NA
    }),
    
    # 3. Average goals for/against last 3 months (90 days)
    last_3months_goals_for_avg = map_dbl(date, ~ {
      matches_last_90d = goals_for[between(date, .x - 90, .x - 1)]
      if(length(matches_last_90d) > 0) mean(matches_last_90d) else NA
    }),
    last_3months_goals_against_avg = map_dbl(date, ~ {
      matches_last_90d = goals_against[between(date, .x - 90, .x - 1)]
      if(length(matches_last_90d) > 0) mean(matches_last_90d) else NA
    }),
    
    # 4. Naive win probability (win share)
    total_matches = row_number(),
    wins = cumsum(home_result == 1),
    draws = cumsum(home_result == 0.5),
    losses = cumsum(home_result == 0),
    win_share = wins / total_matches,
    win_prob_naive = (wins + 0.5 * draws) / total_matches,  # draws as half-win
  ) %>%
  ungroup() %>%
  
  # Now create opponent-specific rolling statistics (last 5 matches against SAME opponent)
  group_by(country_name, opponent_name) %>%
  mutate(
    # 5. Last 5 matches against this specific opponent
    last5_vs_opponent_result_avg = rollapplyr(home_result, width = 5, 
                                               FUN = mean, fill = NA, align = "right"),
    last5_vs_opponent_goals_for_avg = rollapplyr(goals_for, width = 5, 
                                                  FUN = mean, fill = NA, align = "right"),
    last5_vs_opponent_goals_against_avg = rollapplyr(goals_against, width = 5, 
                                                      FUN = mean, fill = NA, align = "right"),
    
    # Cumulative head-to-head record
    h2h_matches = row_number(),
    h2h_wins = cumsum(home_result == 1),
    h2h_draws = cumsum(home_result == 0.5),
    h2h_losses = cumsum(home_result == 0),
    h2h_win_rate = h2h_wins / h2h_matches,
    h2h_points_per_game = (h2h_wins * 3 + h2h_draws * 1) / h2h_matches
  ) %>%
  ungroup() %>%
  
  # Now create SAME variables for the OPPONENT (by joining/merging)
  # First, create opponent's perspective dataset
  left_join(
    monadic_matches %>%
      select(date, country_name, opponent_name, home_result, goals_for, goals_against) %>%
      rename(
        opponent_goals_for = goals_for,
        opponent_goals_against = goals_against,
        opponent_home_result = home_result
      ) %>%
      group_by(opponent_name) %>%
      arrange(opponent_name, date) %>%
      mutate(
        opponent_last5_result_avg = rollapplyr(opponent_home_result, width = 5, 
                                                FUN = mean, fill = NA, align = "right"),
        opponent_last_year_goals_for_avg = map_dbl(date, ~ {
          matches_last_year = opponent_goals_for[between(date, .x - 365, .x - 1)]
          if(length(matches_last_year) > 0) mean(matches_last_year) else NA
        }),
        opponent_last_year_goals_against_avg = map_dbl(date, ~ {
          matches_last_year = opponent_goals_against[between(date, .x - 365, .x - 1)]
          if(length(matches_last_year) > 0) mean(matches_last_year) else NA
        }),
        opponent_win_share = cumsum(opponent_home_result == 1) / row_number()
      ) %>%
      ungroup() %>%
      select(date, country_name = opponent_name, opponent_name = country_name, 
             starts_with("opponent_")),
    by = c("date", "country_name", "opponent_name")
  ) 
```

## Geopolitical data
```{r, echo = F, message = F, warning = F}
#qog <- read.csv("C:/Users/Simon Labarthe/OneDrive/Documents/UNI/ETH/26a/Method III/Project/qog_std_ts_jan26.csv") %>% 

qog <- read.csv("C:/Users/Simon/OneDrive/Documents/UNI/ETH/26a/Method III/Project/qog_std_ts_jan26.csv") %>% 
  rename(country_name = cname) %>% 
  select(
    country_name,
    year,

    bicc_milper,        # Military Personnel Index 
    bicc_milexp,        # Military Expenditure Index 
    wdi_expmil,         # Military expenditure % of GDP 

    wdi_pop,            # Total Population
    wdi_popurb,         # Urban Population

    wdi_birth,          # Birth Rate (per 1,000)
    wdi_death,          # Death Rate (per 1,000)
    wdi_mortinf,        # Infant Mortality (per 1,000 live births)
    wdi_lifexp,         # Life Expectancy at birth

    wpp_medianage,      # Median Age as of 1 July (WPP) 
    wpp_sexratio,       # Sex Ratio: males per 100 females (WPP) 


    wdi_export,         # Exports of goods & services (% of GDP)
    wdi_import,         # Imports of goods & services (% of GDP)
    gle_exp,          # Total Exports in millions current USD (Gleditsch)
    gle_imp,          # Total Imports in millions current USD (Gleditsch)

    wdi_area,           # Land area in sq. km 

    wdi_eneuse,         # Energy use (kg of oil equivalent per capita)

   vdem_polyarchy,   # Democracy level

    ucdp_type3        # Number of internal armed conflicts 
  )
```

## Final dataset
```{r, echo = F, message = F, warning = F}
# First, join QoG data for BOTH home and away teams
data <- monadic_data %>%
  left_join(qog, by = c("country_name", "year")) %>%
  left_join(
    qog %>% 
      rename_with(~paste0("opponent_", .x), -c(country_name, year)),
    by = c("opponent_name" = "country_name", "year")
  ) %>%
  
  # Create meaningful differences and ratios
  mutate(
    # MILITARY & POWER (subtraction makes sense - who has more military might)
    milper_diff = bicc_milper - opponent_bicc_milper,
    milexp_diff = bicc_milexp - opponent_bicc_milexp,
    milexp_gdp_diff = wdi_expmil - opponent_wdi_expmil,
    
    # POPULATION (ratio makes more sense - log ratio for interpretation)
    pop_ratio = log(wdi_pop / opponent_wdi_pop),  # log because 100M vs 10M = 2.3 vs 0.23
    popurb_ratio = log(wdi_popurb / opponent_wdi_popurb),
    popurb_pct_diff = (wdi_popurb / wdi_pop) - (opponent_wdi_popurb / opponent_wdi_pop),
    
    # DEMOGRAPHICS (subtraction for rates, ratio for ages)
    birth_rate_diff = wdi_birth - opponent_wdi_birth,  # per 1000
    death_rate_diff = wdi_death - opponent_wdi_death,
    infant_mort_diff = wdi_mortinf - opponent_wdi_mortinf,
    life_exp_diff = wdi_lifexp - opponent_wdi_lifexp,  # positive = home lives longer
    median_age_diff = wpp_medianage - opponent_wpp_medianage,  # positive = home older
    sex_ratio_diff = wpp_sexratio - opponent_wpp_sexratio,
    
    # ECONOMY (log ratios for scale)
    export_gdp_diff = wdi_export - opponent_wdi_export,  # % of GDP difference
    import_gdp_diff = wdi_import - opponent_wdi_import,
    total_export_ratio = log(gle_exp / opponent_gle_exp),  # millions USD
    total_import_ratio = log(gle_imp / opponent_gle_imp),
    trade_openness_diff = (wdi_export + wdi_import) - (opponent_wdi_export + opponent_wdi_import),
    
    # GEOGRAPHY & ENERGY
    area_ratio = log(wdi_area / opponent_wdi_area),
    energy_use_diff = wdi_eneuse - opponent_wdi_eneuse,  # kg oil equivalent per capita
    
    # POLITICS (difference in democracy)
    democracy_diff = vdem_polyarchy - opponent_vdem_polyarchy,  # positive = home more democratic
    
    # CONFLICT (difference in internal conflicts)
    conflict_diff = ucdp_type3 - opponent_ucdp_type3,  # positive = home more conflict-prone
    conflict_any = factor(ucdp_type3 > 0 | opponent_ucdp_type3 > 0, 
                          levels = c(FALSE, TRUE), 
                          labels = c("No conflict", "At least one has conflict")),
    
    # INTERACTION: Home advantage in democracy + development
    home_advantage_geo = (home == 1) * (
      (wdi_lifexp - opponent_wdi_lifexp) / 10 +  # life expectancy advantage (decades)
      (wdi_pop - opponent_wdi_pop) / 1e7 +       # population advantage (10M units)
      (wdi_expmil - opponent_wdi_expmil) / 5     # military spending advantage (% GDP)
    )
  ) %>%
  
  # Flag missing geopolitical data (common for smaller nations)
  mutate(
    geo_data_complete = !is.na(milper_diff) & !is.na(pop_ratio),
    geo_any_missing = is.na(milper_diff) | is.na(pop_ratio)
  )
```


